---
title: "Vocabulary ICC and power for the border cohort"
output:
  html_document:
    toc: true
    toc_float: true
---

This document estimates how much of the variation in LEVANTE Vocabulary scores
is stable between children of the same age (the ICC), and uses it to calculate
power for a three-wave longitudinal study relating an environmental index to
the growth of Vocabulary scores.

What comes from data and what is assumed:

| Quantity | Source |
|---|---|
| ICC of Vocabulary scores at a fixed age | Estimated here, from LEVANTE pilot data (Leipzig) |
| Within-age SD of Vocabulary scores | Estimated here |
| Average yearly growth in Vocabulary scores | Estimated here |
| Target effect, β = .08 SD/year | Chosen (proposal) |
| Between-child SD of growth rates, .15 SD/year | Assumed |
| Three annual waves, 20% attrition by wave 3, α = .05 | Design assumptions |

```{r setup, message = FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(lme4)
library(simr)

# fix the random seed so bootstrap intervals are identical on every run
seed <- 20260914
set.seed(seed)
cat("Random seed:", seed, "\n")
```

# Data

## Download

The scores table comes from the public LEVANTE pilot release on Redivis
(`levante_data_pilots`, current version). Access needs a Redivis account, the
signed LEVANTE data use agreement, and an API token with the `data.data` scope
saved in `~/.Renviron` as `REDIVIS_API_TOKEN`. This chunk only runs if
`data/scores.csv` is not already saved.

```{r download, eval = !file.exists("data/scores.csv")}
# qualified Redivis reference: owner.dataset:id:version.table:id
table_ref <- "levante.levante_data_pilots:68kn:current.scores:pgms"
url <- paste0("https://redivis.com/api/v1/tables/", table_ref, "/rows?format=csv")

token <- Sys.getenv("REDIVIS_API_TOKEN")
if (token == "") stop("Set REDIVIS_API_TOKEN in ~/.Renviron and restart R.")

# build the request, attach the token, and write the response straight to a CSV file
dir.create("data", showWarnings = FALSE)
httr2::request(url) |>
  httr2::req_headers(Authorization = paste("Bearer", token)) |>
  httr2::req_perform(path = "data/scores.csv")
```

## Load and filter

```{r load}
# read the full scores table (all tasks, all pilot sites)
scores <- read_csv("data/scores.csv", show_col_types = FALSE)

# record which data this document was run on
cat("File downloaded:", format(file.mtime("data/scores.csv"), "%Y-%m-%d"), "\n")
cat("Rows in scores table:", nrow(scores), "\n")
```

```{r filter}
# keep Vocabulary runs, then runs without an exclusion flag, then runs with a score
vocab_all    <- filter(scores, task_id == "vocab")
vocab_scored <- filter(vocab_all, is.na(exclusion))
vocab        <- filter(vocab_scored, !is.na(score))

# how many runs survive each step
tibble(step = c("Vocabulary runs", "no exclusion flag", "has a score"),
       runs = c(nrow(vocab_all), nrow(vocab_scored), nrow(vocab)))

# reasons runs were excluded
count(vocab_all, exclusion)
```

## Which sites have children tested more than once?

```{r sites}
# one row per child: number of scores and age at first test
kids <- vocab |>
  group_by(dataset, user_id) |>
  summarise(n_scores = n(), first_age = min(age), .groups = "drop")

# per site: children, retested children, missing ages, and age range at first test
kids |>
  group_by(dataset) |>
  summarise(children = n(),
            retested = sum(n_scores >= 2),
            missing_age = sum(is.na(first_age)),
            youngest = round(min(first_age, na.rm = TRUE), 1),
            oldest = round(max(first_age, na.rm = TRUE), 1))
```

Only Leipzig (`pilot_mpieva_de_main`) has enough retested children with
recorded ages, so all estimates below use Leipzig.

## The Leipzig sample

```{r leipzig}
# Leipzig Vocabulary scores with a recorded age, ordered by child and then by age
leipzig <- vocab |>
  filter(dataset == "pilot_mpieva_de_main", !is.na(age)) |>
  arrange(user_id, age)

# summarise within each child first
leipzig_kids <- leipzig |>
  group_by(user_id) |>
  summarise(n_scores = n(), first_age = min(age), .groups = "drop")

# then describe the sample across children
leipzig_kids |>
  summarise(children = n(),
            scores = sum(n_scores),
            one_score = sum(n_scores == 1),
            two_scores = sum(n_scores == 2),
            three_or_more = sum(n_scores >= 3),
            youngest = round(min(first_age), 1),
            median_age = round(median(first_age), 1),
            oldest = round(max(first_age), 1))

# scoring model and test version behind these scores
count(leipzig, scoring_model, registry_version, score_type, language)
count(leipzig, adaptive)
```

`score` is an IRT ability estimate from a Rasch model fitted to German-language
Vocabulary responses, in logits.

## Retest pairs and a scoring check

LEVANTE Vocabulary scores were affected by a scoring bug, since fixed. For
Leipzig retest pairs, the LEVANTE longitudinal analyses report a slope of T2
on T1 of .15 with the buggy scores and .62 with corrected scores, and mean
change of −.91 vs .00 logits. Values near the corrected ones confirm these
are corrected scores.

```{r pairs}
# one row per retested child: first and second score, versions, and the gap between tests
pairs <- leipzig |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  summarise(age_t1 = age[1], age_t2 = age[2],
            t1 = score[1], t2 = score[2],
            adaptive_t1 = adaptive[1], adaptive_t2 = adaptive[2],
            gap_months = 12 * (age[2] - age[1]),
            .groups = "drop")

# retest summaries across children (each child contributes one pair)
pairs |>
  summarise(children = n(),
            median_gap_months = median(gap_months),
            min_gap_months = min(gap_months),
            max_gap_months = max(gap_months),
            retest_r = cor(t1, t2),
            slope_t2_on_t1 = coef(lm(t2 ~ t1))[2],
            mean_change_logits = mean(t2 - t1)) |>
  print(width = Inf)

# test version at first and second test
count(pairs, adaptive_t1, adaptive_t2)
```

```{r pairs-plot, fig.width = 6.5, fig.height = 4.5}
# first vs second score for each retested child; dashed line = identical scores
ggplot(pairs, aes(t1, t2, colour = age_t1)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  labs(x = "Vocabulary score at first test (IRT ability, logits)",
       y = "Vocabulary score at second test (IRT ability, logits)",
       colour = "Age at first test (years)")
```

```{r age-plot, fig.width = 6.5, fig.height = 4.5}
# every Leipzig score against age, with the smooth age curve used in the ICC model
ggplot(leipzig, aes(age, score)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, df = 3), colour = "black") +
  labs(x = "Age at test (years)",
       y = "Vocabulary score (IRT ability, logits)")
```

# ICC at a fixed age

## Model

$$
\text{score}_{ij} = f(\text{age}_{ij}) + \gamma\,\text{adaptive}_{ij} + u_i + e_{ij},
\qquad u_i \sim N(0, \tau^2), \quad e_{ij} \sim N(0, \sigma^2)
$$

for child *i* at test *j*. $f$ is a natural spline in age with 3 degrees of
freedom, which removes the average age trend. `adaptive` shifts scores from the
adaptive version of the task. $u_i$ is each child's stable deviation from
same-age peers; $e_{ij}$ is everything that differs from test to test
(measurement error plus real fluctuation). The ICC is

$$
\text{ICC} = \frac{\tau^2}{\tau^2 + \sigma^2}.
$$

Children with one score contribute to the age curve and to the total
variance; children with two scores separate $\tau^2$ from $\sigma^2$.

```{r icc-model}
# fit the model by REML
icc_model <- lmer(score ~ splines::ns(age, df = 3) + adaptive + (1 | user_id),
                  data = leipzig, REML = TRUE)
summary(icc_model)
```

```{r icc-components}
# pull the two variance components out of the fitted model
variance_components <- as.data.frame(VarCorr(icc_model))
child_var    <- variance_components$vcov[variance_components$grp == "user_id"]
residual_var <- variance_components$vcov[variance_components$grp == "Residual"]

# ICC and the SD of scores among children of the same age
icc_estimate  <- child_var / (child_var + residual_var)
within_age_sd <- sqrt(child_var + residual_var)

tibble(child_variance = child_var,
       residual_variance = residual_var,
       icc = icc_estimate,
       within_age_sd_logits = within_age_sd)
```

## How much of the residual is measurement error?

Each score has a standard error from the IRT model. Its square is the
measurement error variance of that score.

```{r measurement-error}
# average squared standard error within each child, then across children
error_var <- leipzig |>
  group_by(user_id) |>
  summarise(child_mean_se2 = mean(score_se^2), .groups = "drop") |>
  summarise(mean_se2 = mean(child_mean_se2)) |>
  pull(mean_se2)

tibble(residual_variance = residual_var,
       measurement_error_variance = error_var,
       share_of_residual = error_var / residual_var)
```

What this means, in plain terms. Each Vocabulary score comes with a standard
error from the IRT model: how precisely that one test pins down the child's
ability. Squared and averaged, it says how much scores would bounce around from
test to test from imprecision alone (`r round(error_var, 3)`). That is about the
same size as the test-to-test variance in the ICC model (`r round(residual_var, 3)`).

So when a child's score changes between two tests a few months apart, the change
is about what imprecision alone would produce. There is no sign that children's
vocabulary, relative to same-age peers, actually shifts over those months. The two
numbers come from different calculations and are both approximate, so their ratio
being a little above 1 means "about equal", nothing more.

The practical consequence: children's relative vocabulary level is stable, but one
score measures it imprecisely. Among children of the same age, a single score's
reliability is roughly the ICC (`r round(icc_estimate, 2)`). This imprecision is
already built into the ICC, and so into the power calculations below. It also means
an observed association with the environmental index will be weaker than the true
one, which is a reason to consider averaging several language tasks into one outcome.

## Bootstrap confidence intervals

Parametric bootstrap: simulate 500 new datasets from the fitted model, refit
each, recompute the ICC, and take the 2.5th and 97.5th percentiles. This is
done for all ages and for children aged 9 and under at the time of testing.

```{r icc-functions}
# fit the ICC model to any subset of the data
fit_icc <- function(d) {
  lmer(score ~ splines::ns(age, df = 3) + adaptive + (1 | user_id), data = d, REML = TRUE)
}

# ICC from a fitted model: child variance / (child variance + residual variance)
icc_from_model <- function(m) {
  v <- as.data.frame(VarCorr(m))
  v$vcov[v$grp == "user_id"] / sum(v$vcov)
}

# point estimate plus a parametric bootstrap interval; the seed is reset so each call is reproducible
icc_with_ci <- function(d, label, n_boot = 500) {
  m <- fit_icc(d)
  set.seed(seed)
  boot <- bootMer(m, icc_from_model, nsim = n_boot, type = "parametric")
  tibble(ages = label,
         children = n_distinct(d$user_id),
         retested = sum(table(d$user_id) >= 2),
         icc = icc_from_model(m),
         ci_low = unname(quantile(boot$t, 0.025)),
         ci_high = unname(quantile(boot$t, 0.975)))
}
```

```{r icc-bootstrap, warning = FALSE}
# ICC with intervals, all ages and ages 9 and under
icc_results <- bind_rows(
  icc_with_ci(leipzig, "all ages"),
  icc_with_ci(filter(leipzig, age <= 9), "ages 9 and under")
)
icc_results
```

## Does the age range change the ICC?

Without age in the model, older children outscore younger ones at every test,
so age differences look like stable child differences and inflate the ICC.
With age in the model, the ICC should not depend on the age range.

```{r age-range}
# ICC from a model with a given formula, fitted to a given subset
simple_icc <- function(d, formula) icc_from_model(lmer(formula, data = d, REML = TRUE))

no_age_formula     <- score ~ adaptive + (1 | user_id)
spline_age_formula <- score ~ splines::ns(age, df = 3) + adaptive + (1 | user_id)

# attach each child's age at first test, used to form age bands
leipzig_banded <- left_join(leipzig, select(leipzig_kids, user_id, first_age), by = "user_id")
under_9 <- filter(leipzig, age <= 9)

tibble(
  sample = c("all ages", "all ages", "ages 9 and under", "ages 9 and under"),
  age_in_model = c("no", "yes", "no", "yes"),
  age_sd_years = c(sd(leipzig$age), sd(leipzig$age), sd(under_9$age), sd(under_9$age)),
  icc = c(simple_icc(leipzig, no_age_formula), simple_icc(leipzig, spline_age_formula),
          simple_icc(under_9, no_age_formula), simple_icc(under_9, spline_age_formula))
)

# within two-year bands of age at first test, with a linear age term
band_formula <- score ~ age + adaptive + (1 | user_id)
bands <- list("5 to 7" = c(0, 7), "7 to 9" = c(7, 9), "9 to 11" = c(9, 11), "11 and over" = c(11, Inf))

# for each band: keep children whose first test falls in it, then fit the model
bind_rows(lapply(names(bands), function(b) {
  d <- filter(leipzig_banded, first_age >= bands[[b]][1], first_age < bands[[b]][2])
  tibble(band = b,
         children = n_distinct(d$user_id),
         retested = sum(table(d$user_id) >= 2),
         icc = simple_icc(d, band_formula))
}))
```

The bands have only 30–50 retested children each, so their ICCs are noisy.

# Does the retest gap matter?

The Leipzig children were retested a median of 3.6 months apart, but the border
study would test children a year apart. If children's standing drifts over a
year in ways a straight growth line does not capture, that drift counts as
test-to-test noise in an annual study, and the ICC estimated from short gaps
would be too high.

Two pilot sites differ in their retest gaps: Leipzig at 3.6 months and Bogotá at
15.6 months. Comparing them across all tasks with enough retested children gives
a rough sense of how much the gap matters. The comparison is imperfect: site,
language, sample and testing conditions differ alongside the gap.

```{r gap-comparison}
# all tasks at all sites, keeping runs that were scored and have an age
all_tasks <- read_csv("data/scores.csv", show_col_types = FALSE) |>
  filter(is.na(exclusion), !is.na(score), !is.na(age), abs(score) < 8)

# site and task combinations with at least 30 retested children
cells <- all_tasks |>
  group_by(dataset, task_id) |>
  summarise(children = n_distinct(user_id), retested = sum(table(user_id) >= 2), .groups = "drop") |>
  filter(retested >= 30)

# fit the same kind of ICC model in each cell and record the median retest gap:
# lapply walks through the cells, fits the model, and returns one row per cell;
# bind_rows stacks those rows into a table
gap_table <- bind_rows(lapply(seq_len(nrow(cells)), function(i) {

  d <- filter(all_tasks, dataset == cells$dataset[i], task_id == cells$task_id[i])

  # the gap between the first two tests for each retested child
  gaps <- d |>
    group_by(user_id) |>
    filter(n() >= 2) |>
    arrange(age, .by_group = TRUE) |>
    summarise(gap = 12 * (age[2] - age[1]), .groups = "drop")

  # include the version term only where both versions appear in the cell
  f <- if (n_distinct(d$adaptive) > 1) {
    score ~ splines::ns(age, df = 3) + adaptive + (1 | user_id)
  } else {
    score ~ splines::ns(age, df = 3) + (1 | user_id)
  }

  model <- lmer(f, data = d)

  tibble(site = gsub("^pilot_|_main$", "", cells$dataset[i]),
         task = cells$task_id[i],
         retested = cells$retested[i],
         gap_months = round(median(gaps$gap), 1),
         icc = round(icc_from_model(model), 3))
}))

# show the cells, shortest gaps first
gap_table |> arrange(gap_months, site, task) |> print(n = Inf)

# the typical ICC at each gap length
gap_table |>
  mutate(gap_group = ifelse(gap_months < 9, "short gaps (3-5 months)", "annual gaps (15-17 months)")) |>
  group_by(gap_group) |>
  summarise(cells = n(), median_icc = round(median(icc), 3))
```

```{r gap-paired}
# compare the same task across the two sites, in steps:
# select keeps the three columns needed; pivot_wider puts each site in its own column;
# filter keeps tasks measured at both sites; mutate computes the ratio of the two ICCs
gap_table |>
  select(site, task, icc) |>
  tidyr::pivot_wider(names_from = site, values_from = icc) |>
  filter(!is.na(mpieva_de), !is.na(uniandes_co_bogota)) |>
  mutate(ratio = round(uniandes_co_bogota / mpieva_de, 2)) |>
  rename(leipzig_3.6_months = mpieva_de, bogota_15.6_months = uniandes_co_bogota)
```

Bogotá's Vocabulary cell is too thin to include here (29 retested children). Its
ICC comes out at .12, but that rests on a handful of children whose scores moved
2 to 3 logits, and the retest correlation rises to .55 if the three largest
movers are dropped. Bogotá's wave-1 Vocabulary runs also attempted far fewer
items than Leipzig's (a median of 49 against 115), so those scores are less
precise. It sits well below Bogotá's own better-measured tasks (math .58, mental
rotation .55).

What this suggests: stability at annual gaps is somewhat lower than at three
months, with the typical ICC falling from about .46 to about .35 across tasks,
rather than collapsing. The ICC used above (`r round(icc_estimate, 2)`) is
therefore probably a little optimistic for an annual design, and the lower end of
its confidence interval (`r round(icc_results$ci_low[icc_results$ages == "all ages"], 2)`) is a reasonable
stand-in for the annual case. Because site and gap are confounded here, this
bounds the problem rather than settling it.

# Average growth in Vocabulary scores

Used only to put the effect size in context. Two estimates:

- **Within children**: each retested child's change divided by the time between
  tests, averaged across children. Every retested Leipzig child took the
  non-adaptive version first and the adaptive version second, so this change
  also contains any difference between versions, and short gaps make each
  child's yearly rate noisy.
- **Across ages**: the slope of score on age, using each child's first score.
  This is not affected by the version switch, and it is the estimate used below.

```{r average-growth}
# within children: change per year for each retested child, then summarised across children
within_child_growth <- pairs |>
  mutate(change_per_year = (t2 - t1) / (gap_months / 12)) |>
  summarise(children = n(),
            mean_logits_per_year = mean(change_per_year),
            median_logits_per_year = median(change_per_year))
within_child_growth

# across ages: keep each child's earliest score, then regress score on age
first_scores <- leipzig |>
  group_by(user_id) |>
  slice_min(age, n = 1, with_ties = FALSE) |>
  ungroup()

age_slope_model <- lm(score ~ age + adaptive, data = first_scores)
average_growth_logits <- coef(age_slope_model)[["age"]]
age_slope_ci <- confint(age_slope_model, "age")

tibble(children = nrow(first_scores),
       age_slope_logits_per_year = average_growth_logits,
       ci_low = age_slope_ci[1],
       ci_high = age_slope_ci[2],
       age_slope_sd_per_year = average_growth_logits / within_age_sd)
```

# Power for the border cohort

## Approach

Power is estimated by simulation with the **simr** package (Green & MacLeod,
2016, *Methods in Ecology and Evolution*). The steps:

- Build an artificial version of the study: the children, the waves at which
  each child is tested, who drops out, and each child's environmental index score.
- Write down the analysis model and fill in its values: the ICC estimated
  above, the assumed spread of growth rates, and the growth effect to detect.
- simr generates Vocabulary scores for that study from the model, fits the
  model to them with `lmer`, and tests the index × time effect. It does this
  many times.
- Power is the share of simulated studies in which the index × time effect is
  significant at α = .05.

## The model

For child *i* at wave *j* (time $t_{ij}$ = 0, 1 or 2 years after the first test):

$$
y_{ij} = b_0 + u_{0i} + (b_1 + u_{1i})\,t_{ij} + \beta_{\text{level}} X_i + \beta_{\text{growth}} X_i t_{ij} + e_{ij}
$$

- $y_{ij}$ is the child's Vocabulary score at that wave, in units of the
  within-age SD estimated above (logits divided by that SD, not re-standardized
  at each wave, so growth is kept).
- $b_0$ is the average starting score (set to 0; it does not affect power) and
  $b_1$ is the average yearly growth (the age slope estimated above, in SD per year).
- $u_{0i}$ is how far child *i* starts above or below average; its variance is the ICC.
- $u_{1i}$ is how much faster or slower than average child *i* grows; its SD is
  the assumed growth SD. Starting level and growth rate are assumed uncorrelated.
- $e_{ij}$ is test-to-test noise, with variance 1 − ICC, so a single first-wave
  score has total variance 1.
- $X_i$ is the child's environmental index, standardized (mean 0, SD 1).
- $\beta_{\text{level}}$ is set to 0: it is not the effect being tested, and it
  does not change power for the growth effect.
- $\beta_{\text{growth}}$ is the tested effect: extra growth per year, in SD,
  for each SD of the index.

In `lmer` notation this is `y ~ time * index + (1 + time | child)`.

## Settings

```{r power-settings}
# the three enrolment sizes being compared; the L stores them as whole numbers (integers),
# so they can be matched exactly later without comparing decimal numbers
n_enrolled <- c(300L, 320L, 360L)

# the share of enrolled children still being tested at wave 1, wave 2 and wave 3:
# everyone at wave 1, 90% at wave 2 (10% lost), 80% at wave 3 (20% lost in total)
retention <- c(1.00, 0.90, 0.80)

# how much children differ in their yearly growth rate, as an SD in within-age SD units per year;
# this is an assumption, not estimated from the pilot data
growth_sd <- 0.15

# average yearly growth in within-age SD units: the age slope in logits divided by the within-age SD;
# it sets b1 in the model, which makes the simulated scores realistic but does not change power
average_growth_sd <- average_growth_logits / within_age_sd

# the significance threshold for a two-sided test
alpha <- 0.05

# the power we want the study to have (an 80% chance of detecting a true effect)
target_power <- 0.80

# the growth effect whose power we estimate: 0.08 SD of extra Vocabulary growth per year
# for each 1 SD higher on the environmental index
target_effect <- 0.08

# the number of simulated studies for each setting in the main table; with 1000 simulations
# a power near .75 is estimated to within about ±0.03 (95% interval)
n_sims <- 1000

# fewer simulations for the sensitivity table, which is only there to show the direction of change
n_sims_sensitivity <- 500

# the ICC estimate and its 95% interval from the bootstrap above, stored with names so each can be looked up
icc_all <- filter(icc_results, ages == "all ages")
icc_values <- c(estimate = icc_all$icc, ci_low = icc_all$ci_low, ci_high = icc_all$ci_high)
icc_values
```

## Building the simulated study

```{r power-design}
# make_design() builds the explanatory data for one study with n enrolled children:
# one row per test actually taken, with the child's id, the time of the test in years,
# and the child's environmental index score (no Vocabulary scores yet; simr adds those)
make_design <- function(n) {

  # reset the random seed so the same index scores and dropout pattern are produced every time
  set.seed(seed)

  # how many children are last tested at wave 1, wave 2 and wave 3:
  # -diff(retention) is c(0.10, 0.10), the shares lost after waves 1 and 2, and retention[3] is the
  # 0.80 who complete all three; multiplying by n and rounding turns shares into counts
  n_last_wave <- round(n * c(-diff(retention), retention[3]))

  # stop with an error if rounding made the counts add up to something other than n
  stopifnot(sum(n_last_wave) == n)

  # one row per child:
  # child is an id stored as a factor, because lmer groups rows by it;
  # index is drawn from a normal distribution, then scale() makes its mean exactly 0 and SD exactly 1;
  # last_wave repeats 1, 2 and 3 as many times as the counts above, then sample() shuffles them across children
  children <- tibble(
    child = factor(seq_len(n)),
    index = as.numeric(scale(rnorm(n))),
    last_wave = sample(rep(1:3, times = n_last_wave))
  )

  # expand to one row per test in steps:
  # crossing gives every child a row for each time 0, 1 and 2;
  # filter then drops the waves after each child's last test (time 0 is wave 1, so a child whose
  # last wave is 2 keeps times 0 and 1)
  children |>
    tidyr::crossing(time = 0:2) |>
    filter(time < last_wave)
}

# look at the design for N = 360: how many children are tested at each time point
# (count() tallies the rows at each value of time)
make_design(360L) |>
  count(time, name = "children_tested")
```

```{r power-model}
# make_model() builds an artificial lmer model with the effects and variances filled in,
# for a given design, ICC, growth effect and growth SD
make_model <- function(design, icc, effect = target_effect, slope_sd = growth_sd) {

  # makeLmer() from simr creates a model object without any real Vocabulary data:
  # formula is the analysis model;
  # fixef gives the fixed effects in the order lmer lists them for this formula:
  #   intercept (b0 = 0), time (b1 = average growth), index (beta level = 0), time:index (beta growth);
  # VarCorr is the 2 x 2 covariance matrix of each child's own level and growth:
  #   diagonal entries are the ICC and the growth SD squared, off-diagonal entries 0 (uncorrelated);
  # sigma is the SD of test-to-test noise, the square root of 1 - ICC;
  # data is the design built by make_design()
  makeLmer(y ~ time * index + (1 + time | child),
           fixef = c(0, average_growth_sd, 0, effect),
           VarCorr = diag(c(icc, slope_sd^2)),
           sigma = sqrt(1 - icc),
           data = design)
}

# build the model for N = 360 at the ICC estimate and print what was filled in, to check it
example_model <- make_model(make_design(360L), icc_values[["estimate"]])
fixef(example_model)
VarCorr(example_model)
sigma(example_model)
```

```{r power-run-function}
# run_power() simulates many studies for one setting and returns the estimated power
run_power <- function(n, icc, slope_sd = growth_sd, nsim = n_sims) {

  # build the study design and the model for this setting
  model <- make_model(make_design(n), icc, slope_sd = slope_sd)

  # powerSim() repeats nsim times: simulate Vocabulary scores from the model, refit
  # y ~ time * index + (1 + time | child) with lmer, and test the time:index effect;
  # fixed("time:index", "sa") picks that effect and the Satterthwaite t-test (from lmerTest);
  # alpha is the significance threshold; seed makes the simulations reproducible;
  # progress = FALSE hides the progress bar
  sims <- powerSim(model,
                   test = fixed("time:index", "sa"),
                   nsim = nsim,
                   alpha = alpha,
                   seed = seed,
                   progress = FALSE)

  # summary() gives the number of significant results, the number of simulations,
  # the power (their ratio), and a 95% interval for the power
  result <- summary(sims)

  # return one row: the setting, the power with its interval, and how many simulated fits
  # produced a warning (usually a singular fit, when the growth-rate variance is estimated at 0)
  # or an error
  tibble(n = n,
         icc = icc,
         growth_sd = slope_sd,
         simulations = nsim,
         power = result$mean,
         power_ci_low = result$lower,
         power_ci_high = result$upper,
         fits_with_warnings = length(unique(sims$warnings$index)),
         fits_with_errors = length(unique(sims$errors$index)))
}
```

## Power at each sample size

This chunk runs `r 3 * 3` settings × `r n_sims` simulations and takes several
minutes. It is cached, so it only reruns when its code or inputs change.

```{r power-table, cache = TRUE, message = FALSE, warning = FALSE, cache.extra = list(icc_values, n_enrolled, retention, growth_sd, average_growth_sd, target_effect, n_sims, seed)}
# every combination of sample size and ICC value (3 x 3 = 9 settings), in steps:
# expand.grid makes the combinations; mutate looks up the numeric ICC for each name
settings <- expand.grid(n = n_enrolled, icc_used = names(icc_values), stringsAsFactors = FALSE) |>
  mutate(icc = icc_values[icc_used])

# run the simulations for each setting in turn:
# lapply goes through the row numbers of settings, runs run_power() for that row,
# and adds a column saying which ICC value was used; bind_rows stacks the nine results into one table
power_table <- bind_rows(lapply(seq_len(nrow(settings)), function(i) {
  run_power(settings$n[i], settings$icc[i]) |>
    mutate(icc_used = settings$icc_used[i], .before = icc)
}))
```

The smallest effect detectable with 80% power is worked out from the simulated
power. For a test like this, power ≈ Φ(β / SE − 1.96), where Φ is the normal
cumulative distribution. Solving for the standard error gives
SE = β / (Φ⁻¹(power) + 1.96), and the effect needed for 80% power is
SE × (1.96 + 0.84).

```{r power-smallest-effect}
# add the smallest detectable effect to the table in steps:
# implied_se is the standard error implied by the simulated power at the target effect;
# smallest_effect_sd_per_year is that standard error times 1.96 + 0.84 (the z values for a two-sided
# .05 test and for 80% power)
power_table <- power_table |>
  mutate(implied_se = target_effect / (qnorm(power) + qnorm(1 - alpha / 2)),
         smallest_effect_sd_per_year = implied_se * (qnorm(1 - alpha / 2) + qnorm(target_power)))

# show every column
print(power_table, width = Inf)
```

```{r power-plot, fig.width = 7, fig.height = 4.5}
# readable names for the three ICC values, used in the legend and the labels
icc_labels <- c(estimate = sprintf("ICC estimate (%.2f)", icc_values[["estimate"]]),
                ci_low   = sprintf("ICC 95%% CI low (%.2f)", icc_values[["ci_low"]]),
                ci_high  = sprintf("ICC 95%% CI high (%.2f)", icc_values[["ci_high"]]))

# prepare the plot data: add the readable ICC name, ordered low, estimate, high
plot_data <- power_table |>
  mutate(icc_label = factor(icc_labels[icc_used],
                            levels = icc_labels[c("ci_low", "estimate", "ci_high")]))

# draw the plot layer by layer:
# a dashed horizontal line at 80% power;
# a line and points for each ICC value, shifted slightly sideways so the error bars do not overlap;
# vertical error bars showing the 95% interval of each simulated power;
# a label at the right-hand end of each line naming the ICC value;
# fixed colours, the y axis from 0.4 to 1, extra room on the right for the labels, and axis labels with units
dodge <- position_dodge(width = 6)
ggplot(plot_data, aes(n, power, colour = icc_label, group = icc_label)) +
  geom_hline(yintercept = target_power, linetype = "dashed", colour = "grey40") +
  geom_line(position = dodge) +
  geom_point(position = dodge, size = 2.5) +
  geom_errorbar(aes(ymin = power_ci_low, ymax = power_ci_high), width = 0, position = dodge) +
  geom_text(data = filter(plot_data, n == max(n_enrolled)),
            aes(label = icc_label), hjust = 0, nudge_x = 8, size = 3.2, show.legend = FALSE) +
  scale_colour_manual(values = c("#eb6834", "#2a78d6", "#1baf7a")) +
  scale_x_continuous(breaks = n_enrolled, expand = expansion(mult = c(0.05, 0.45))) +
  scale_y_continuous(limits = c(0.4, 1)) +
  labs(x = "Children enrolled at wave 1 (count)",
       y = "Power to detect growth effect of .08 SD/year (probability)",
       colour = NULL,
       caption = "Points: simulated power (1000 simulations each). Bars: 95% interval. Dashed line: 80% power.") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Sensitivity to the assumed growth SD

The between-child SD of growth rates cannot be estimated from these data: no
child has more than two scores, and those are a few months apart. This table
shows how much the answer depends on that assumption, at N = 360 and the ICC
estimate, with `r n_sims_sensitivity` simulations each.

```{r growth-sd-sensitivity, cache = TRUE, message = FALSE, warning = FALSE, cache.extra = list(icc_values, retention, average_growth_sd, target_effect, n_sims_sensitivity, seed)}
# run the simulations once for each growth SD to try, then stack the results into one table:
# lapply goes through the growth SDs and calls run_power() at N = 360 and the ICC estimate
growth_sd_table <- bind_rows(lapply(c(0.05, 0.10, 0.15, 0.25), function(sd_value) {
  run_power(360L, icc_values[["estimate"]], slope_sd = sd_value, nsim = n_sims_sensitivity)
}))

# show every column
print(growth_sd_table, width = Inf)
```

More spread in growth rates makes the average growth rate harder to pin down,
so power falls, but only modestly across this range.

# What the effect sizes mean

Effects here are in SD of Vocabulary score per year, for each SD of the
environmental index. That is a difference in growth *rates*, so it accumulates:
the gap between two children keeps widening as long as the study runs. The
function below restates any such effect in easier terms.

```{r effect-conversions}
# describe_effect() restates a growth effect (in SD per year, per SD of the index) in other units
describe_effect <- function(beta) {

  # the same effect in logits per year: SD units times the within-age SD of Vocabulary scores
  logits_per_year <- beta * within_age_sd

  # how many months of ordinary vocabulary development that equals: the effect divided by
  # how much scores rise with one year of age, times 12 months
  months_per_year <- 12 * logits_per_year / average_growth_logits

  # the gap between two children 1 SD apart on the index after two years (wave 1 to wave 3)
  gap_by_wave_3 <- 2 * beta

  # the SD of scores at wave 3: baseline variance 1, plus the spread added by two years of
  # differing growth rates, plus the spread added by the index effect itself
  wave_3_sd <- sqrt(1 + 4 * growth_sd^2 + 4 * beta^2)

  tibble(
    # the effect as given
    effect_sd_per_year = beta,
    # extra months of vocabulary growth per year, for a child 1 SD higher on the index
    extra_months_per_year = months_per_year,
    # the same, accumulated over the two years from wave 1 to wave 3
    extra_months_by_wave_3 = 2 * months_per_year,
    # Cohen's d at wave 3 for two children 1 SD apart on the index
    cohens_d_1_sd_apart = gap_by_wave_3 / wave_3_sd,
    # Cohen's d at wave 3 for the top half vs the bottom half of the index: the average index values
    # of the two halves differ by 2 * sqrt(2 / pi), about 1.6 SD
    cohens_d_median_split = 2 * sqrt(2 / pi) * gap_by_wave_3 / wave_3_sd,
    # the correlation between the index and children's true growth rates
    r_with_true_growth = beta / sqrt(beta^2 + growth_sd^2)
  )
}
```

## The effect we power for

```{r target-effect-meaning}
# restate the target effect of .08 SD/year in the other units
describe_effect(target_effect) |> print(width = Inf)
```

Read the row like this: a child 1 SD higher on the environmental index gains
about that many extra months of vocabulary development per year, and roughly
twice that by the third wave. The Cohen's d columns compare children's scores
at wave 3: first for two children 1 SD apart on the index, then for the top
half against the bottom half of the index. By Cohen's rule of thumb, 0.2 is
small.

## The smallest effects each sample size can detect

```{r smallest-effect-meaning}
# restate each sample size's smallest detectable effect in the same units, in steps:
# filter keeps the rows simulated at the ICC estimate (one per sample size);
# lapply runs describe_effect() on each of those effects and bind_rows stacks the results;
# the sample sizes are added back as the first column
smallest_effects <- power_table |>
  filter(icc_used == "estimate")

bind_rows(lapply(smallest_effects$smallest_effect_sd_per_year, describe_effect)) |>
  mutate(n = smallest_effects$n, .before = everything()) |>
  print(width = Inf)
```

## Why these effects look small in SD but not in correlation

Children's growth rates do not vary much: the assumed SD is
`r growth_sd` SD/year. So an effect of `r target_effect` SD/year explains a
sizeable share of the differences in how fast children learn vocabulary, which
is why the correlation with true growth rates is around .5 even though Cohen's d
is small. The d values are small because they compare scores at a single wave,
where stable differences between children, which are much larger, dominate.

# Reproducibility check

Re-run the all-ages bootstrap with the same seed and confirm the result matches.

```{r reproducibility, warning = FALSE}
# repeat the bootstrap and compare with the first run, allowing only floating-point tolerance
second_run <- icc_with_ci(leipzig, "all ages")
identical_results <- isTRUE(all.equal(icc_all, second_run, tolerance = 1e-10))
cat("Bootstrap results identical on re-run:", identical_results, "\n")
```

Re-run one simr setting (N = 360 at the ICC estimate) with the same seed and
confirm the simulated power matches the table.

```{r reproducibility-simr, message = FALSE, warning = FALSE}
# simulate the N = 360, ICC estimate setting again
repeat_run <- run_power(360L, icc_values[["estimate"]])

# the same setting from the power table
first_run <- filter(power_table, n == 360L, icc_used == "estimate")

# compare the two powers, allowing only floating-point tolerance
cat("simr power identical on re-run:",
    isTRUE(all.equal(first_run$power, repeat_run$power, tolerance = 1e-12)), "\n")
```

# Methods summary

```{r methods-numbers, include = FALSE}
# numbers used in the text below
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
n_children   <- nrow(leipzig_kids)
n_scores     <- sum(leipzig_kids$n_scores)
n_retested   <- sum(leipzig_kids$n_scores >= 2)
gap_median   <- median(pairs$gap_months)
row_360      <- filter(power_table, n == 360L, icc_used == "estimate")
row_300      <- filter(power_table, n == 300L, icc_used == "estimate")
```

We estimated the between-child share of variance (ICC) in LEVANTE Vocabulary
scores among children of the same age using the public LEVANTE pilot release.
Leipzig was the only site with enough children tested twice: `r n_scores` scores
from `r n_children` children aged `r fmt(min(leipzig$age), 1)`–`r fmt(max(leipzig$age), 1)` years,
`r n_retested` of whom were retested a median of `r fmt(gap_median, 1)` months apart.
Scores were IRT ability estimates (Rasch model, logits). We fitted a linear mixed
model with a natural spline for age (3 df), a fixed effect of adaptive
administration, and a random intercept for child (REML; lme4). The ICC was the
child intercept variance divided by the sum of the intercept and residual
variances, with a 95% interval from 500 parametric bootstrap refits. The ICC was
`r fmt(icc_estimate)` (95% CI `r fmt(icc_all$ci_low)`–`r fmt(icc_all$ci_high)`).

Power was estimated by simulation with simr (Green & MacLeod, 2016). Each
simulated study enrolled N children tested at 0, 1 and 2 years, with 10% last
tested at wave 1 and a further 10% at wave 2, and a standardized environmental
index for each child. Vocabulary scores were generated from a growth model with a
random intercept for each child (variance equal to the ICC), a random slope (SD
`r fmt(growth_sd)` SD/year, uncorrelated with the intercept), residual variance
1 − ICC, and an index × time effect of β = `r fmt(target_effect)` SD/year. Each
simulated dataset was analysed with `lmer(y ~ time * index + (1 + time | child))`,
and the index × time effect was tested with a Satterthwaite t-test (α = `r fmt(alpha)`),
over `r n_sims` simulations per setting. With `r row_360$n` children enrolled, power was
`r fmt(row_360$power)` (95% CI `r fmt(row_360$power_ci_low)`–`r fmt(row_360$power_ci_high)`),
and the smallest effect detectable with 80% power, derived from the simulated power,
was β = `r fmt(row_360$smallest_effect_sd_per_year, 3)` SD/year
(`r row_300$n` children: `r fmt(row_300$smallest_effect_sd_per_year, 3)`).

# Caveats

- One site (Leipzig, German-language scores), ages 5–13. The border cohort is
  4–8 and Spanish/English.
- Retests were months apart, not a year, so the ICC reflects short-term stability.
- Some children switched from the non-adaptive to the adaptive version between
  tests; the model includes a version effect.
- The growth SD (.15 SD/year) and the target effect (.08 SD/year) are assumptions,
  not estimates.
- Simulated power has its own uncertainty: with 1000 simulations, about ±0.03
  (the 95% intervals are shown in the table).

# Session info

```{r session-info}
sessionInfo()
```
